---
title: "Model Building- GAM Models"
output: pdf_document
---

# Load neccessary packages
```{r echo = T, message = FALSE, results='hide', warning=FALSE}
library(tidyverse)
library(tidymodels)
library(ggplot2)
library(purrr)
library(caret)
library(glmnet)
library(mgcv)
library(splines)
library(pdp)
library(earth)
```

# Load in dataset and use 0/1 for binary variables
```{r}
data <- read.csv("data/data_clean.csv") |>
  data.frame() 

data$UnablePregnant <- ifelse(data$UnablePregnant == "Yes", 1,
                              ifelse(data$UnablePregnant == "No", 0, NA))

data$RegularPeriod_12m <- ifelse(data$RegularPeriod_12m == "Yes", 1,
                              ifelse(data$RegularPeriod_12m == "No", 0, NA))

data = data |>
  mutate(
    UnablePregnant = as.factor(UnablePregnant),
    RegularPeriod_12m = as.factor(RegularPeriod_12m)
  )
```

# Model using outome as Age_FirstPeriod (counts)
```{r}
# Model Fitting using mgcv

# Fit model
gam.m1 <- gam(Age_FirstPeriod ~ Age + Race + s(MCNP) + s(MCOP) + s(MECPP) + s(MBP) + 
                s(MCPP) + s(MEP) + s(MEHHP) + s(MEHP) + s(MIBP) + s(MEOHP) + s(MBZP) + 
                Year,
              data = data,
              family = poisson)
summary(gam.m1)
# Generate plots for the nonlinear terms included in the model
plot(gam.m1, pages=4)
```

# Model using outome as Age_FirstPeriod (counts)
```{r eval = FALSE}
# Model Training using caret

# Matrix of predictors
x1 <- model.matrix(Age_FirstPeriod ~ Age + Race + MCNP + MCOP + MECPP + MBP + MCPP + 
                     MEP + MEHHP + MEHP + MIBP + MEOHP + MBZP + Year, data)[, -1]
# Vector of response
y1 <- data$Age_FirstPeriod
# Set up CV
ctrl1 <- trainControl(method = "cv", number = 10)
# Set seed for reproducibility
set.seed(2)
# Train model
gam.fit1 <- train(x1, y1,
                  method = "gam",
                  family = poisson,
                  trControl = ctrl1)
# Best tuning parameters
gam.fit1$bestTune 
# Final model
gam.fit1$finalModel
# Generate plots for the nonlinear terms included in the model
plot(gam.fit1$finalModel, pages=4)
```

# Model using outome as RegularPeriod_12m (binary)
```{r}
# Model Fitting using mgcv

# Fit model
gam.m2 <- gam(RegularPeriod_12m ~ Age + Race + s(MCNP) + s(MCOP) + s(MECPP) + s(MBP) + 
                s(MCPP) + s(MEP) + s(MEHHP) + s(MEHP) + s(MIBP) + s(MEOHP) + s(MBZP) + 
                Year,
              data = data,
              family = binomial)
summary(gam.m2)
# Generate plots for the nonlinear terms included in the model
plot(gam.m2, pages=4)
```

# Model using outome as RegularPeriod_12m (binary)
```{r eval = FALSE}
# Model Training using caret

# Matrix of predictors
x2 <- model.matrix(RegularPeriod_12m ~ Age + Race + MCNP + MCOP + MECPP + MBP + MCPP + 
                    MEP + MEHHP + MEHP + MIBP + MEOHP + MBZP + Year, data)[, -1]
# Vector of response
y2 <- data$RegularPeriod_12m
# Set seed for reproducibility
set.seed(2)
# Train model
gam.fit2 <- train(x2, y2,
                  method = "gam",
                  family = binomial,
                  trControl = ctrl1)
# Best tuning parameters
gam.fit2$bestTune 
# Final model
gam.fit2$finalModel
# Generate plots for the nonlinear terms included in the model
plot(gam.fit2$finalModel, pages=4)
```

# Model using outome as Age_LastPeriod (counts)
```{r}
# Model Fitting using mgcv

# Fit model
gam.m3 <- gam(Age_LastPeriod ~ Age + Race + s(MCNP) + s(MCOP) + s(MECPP) + s(MBP) + 
                s(MCPP) + s(MEP) + s(MEHHP) + s(MEHP) + s(MIBP) + s(MEOHP) + s(MBZP) + 
                Year,
              data = data,
              family = poisson)
summary(gam.m3)
# Generate plots for the nonlinear terms included in the model
plot(gam.m3, pages=4)
```

# Model using outome as Age_LastPeriod (counts)
```{r eval = FALSE}
# Model Training using caret

# Matrix of predictors
x3 <- model.matrix(Age_LastPeriod ~ Age + Race + MCNP + MCOP + MECPP + MBP + MCPP + 
                    MEP + MEHHP + MEHP + MIBP + MEOHP + MBZP + Year, data)[, -1]
# Vector of response
y3 <- data$Age_LastPeriod
# Set seed for reproducibility
set.seed(2)
# Train model
gam.fit3 <- train(x3, y3,
                  method = "gam",
                  family = poisson,
                  trControl = ctrl1)
# Best tuning parameters
gam.fit3$bestTune 
# Final model
gam.fit3$finalModel
# Generate plots for the nonlinear terms included in the model
plot(gam.fit3$finalModel, pages=4)
```

# Model using outome as UnablePregnant (binary)
```{r}
# Model Fitting using mgcv

# Fit model
gam.m4 <- gam(UnablePregnant ~ Age + Race + s(MCNP) + s(MCOP) + s(MECPP) + s(MBP) + 
                s(MCPP) + s(MEP) + s(MEHHP) + s(MEHP) + s(MIBP) + s(MEOHP) + s(MBZP) + 
                Year,
              data = data,
              family = binomial)
summary(gam.m4)
# Generate plots for the nonlinear terms included in the model
plot(gam.m4, pages=4)
```

# Model using outome as UnablePregnant (binary)
```{r eval = FALSE}
# Model Training using caret

# Matrix of predictors
x4 <- model.matrix(UnablePregnant ~ Age + Race + MCNP + MCOP + MECPP + MBP + MCPP + 
                    MEP + MEHHP + MEHP + MIBP + MEOHP + MBZP + Year, data)[, -1]
# Vector of response
y4 <- data$UnablePregnant
# Set seed for reproducibility
set.seed(2)
# Train model
gam.fit4 <- train(x4, y4,
                  method = "gam",
                  family = binomial,
                  trControl = ctrl1)
# Best tuning parameters
gam.fit4$bestTune 
# Final model
gam.fit4$finalModel
# Generate plots for the nonlinear terms included in the model
plot(gam.fit4$finalModel, pages=4)
```



