---
title: "Demographic Data"
output: pdf_document
---

```{r, include = FALSE}
#setup - getting packages, getting raw data ------------------------------------
#loading required packages
library(dplyr)
library(tidyr)
library(janitor)
library(haven)
library(forcats)
library(gtsummary)

#loading raw NHANES demographic data, per year 
#naming convention = "demo" (demographics) year range XX_XX, raw to denote raw files
demo13_14_raw = read_xpt("data/2013_2014/DEMO_H.xpt")
demo15_16_raw = read_xpt("data/2015_2016/DEMO_I.xpt")
demo17_18_raw = read_xpt("data/2017_2018/DEMO_J.xpt")
```

```{r, include = FALSE}
#cleaning data and getting ready to merge, writing function -------------------------
#mapping the list of variables I want to keep
  vars_keep <- c(
  "SEQN", # participant survey ID 
  "RIAGENDR", # gender 
  "RIDAGEYR", # age in years 
  "RIDRETH3", # race/ethnicity 
  "DMDEDUC3", # education - children/youth
  "DMDEDUC2", # education - 20+ 
  "INDHHIN2" # household income 
  )

#this function cleans the data frames and selects only the variables we need, and creates a function that does this for each data frame
clean_demo_df <- function(df, cycle_label) {
  df %>%
    select(any_of(vars_keep)) %>%
    clean_names() %>%
    mutate(
      cycle = cycle_label,
      across(where(is.numeric), ~ na_if(., 9)),
      across(where(is.numeric), ~ na_if(., 77)),
      across(where(is.numeric), ~ na_if(., 99))
    )
}

#applies the function to each of our data frames
demo_13_14_clean <- clean_demo_df(demo13_14_raw, "2013-2014")
demo_15_16_clean <- clean_demo_df(demo15_16_raw, "2015-2016")
demo_17_18_clean <- clean_demo_df(demo17_18_raw, "2017-2018")

demo_all_years <- bind_rows(
  demo_13_14_clean,
  demo_15_16_clean,
  demo_17_18_clean
)

#getting ready for a table - creating a function and changing the variables to their corresponding meaning
clean_label <- function(df, cycle_label) {df %>% 
    mutate(gender = factor(riagendr, levels = c(1, 2), labels = c("Male", "Female")),
    race_eth = factor(ridreth3,levels = c(1, 2, 3, 4, 6, 7),
      labels = c(
        "Mexican American",
        "Other Hispanic",
        "Non-Hispanic White",
        "Non-Hispanic Black",
        "Non-Hispanic Asian",
        "Other / Multi-racial")),
     race_eth = fct_collapse(
      race_eth,
      "Hispanic" = c("Mexican American", "Other Hispanic"),
      "Non-Hispanic White" = "Non-Hispanic White",
      "Non-Hispanic Black" = "Non-Hispanic Black",
      "Non-Hispanic Asian" = "Non-Hispanic Asian",
      "Other / Multi-racial" = "Other / Multi-racial"),
    edu_adult = factor(dmdeduc2, levels = c(1, 2, 3, 4, 5),
      labels = c(
        "< 9th grade",
        "9–11th grade",
        "High school / GED",
        "Some college / AA",
        "College graduate+")),
    dmdeduc2 = case_when(dmdeduc3 %in% c(0) ~ "Never Attended School",
                         dmdeduc3 %in% c(1, 2, 3, 4, 5, 6, 7, 8, 55, 66) ~ ">9th Grade",
                         dmdeduc3 %in% c(9) ~ "9th Grade",
                         dmdeduc3 %in% c(10) ~ "10th Grade",
                         dmdeduc3 %in% c(11) ~ "11th Grade",
                         dmdeduc3 %in% c(12) ~ "12th Grade",
                         dmdeduc3 %in% c(13, 14) ~ "High School Graduate/GED",
                         dmdeduc3 %in% c(15) ~ "More than high school"),
    edu_youth = dmdeduc2,
    
    indhhin2 = case_when(indhhin2 %in% c(1, 2, 3, 4, 5, 13) ~ "<25,000",
                         indhhin2 %in% c(6, 7) ~ "25,000 - 44,999",
                         indhhin2 %in% c(8, 9, 10) ~ "50,000 - 74,999",
                         indhhin2 %in% c(14, 15) ~ ">75,000"),
    income = indhhin2,
    age = ridageyr,
    cycle = cycle_label) %>%
  select(-indhhin2, -dmdeduc2, -riagendr, -ridageyr, -dmdeduc3, -ridreth3) %>%
    filter(gender == "Female")
}

#now running the data frames through this 
demo_13_14_label <- clean_label(demo_13_14_clean, "2013-2014")
demo_15_16_label <- clean_label(demo_15_16_clean, "2015-2016")
demo_17_18_label <- clean_label(demo_17_18_clean, "2017-2018")
```

```{r, include = FALSE}
#all years
demo_all_table = bind_rows (demo_13_14_label, demo_15_16_label, demo_17_18_label)

demo_all_table %>%
  select(gender, race_eth, edu_adult, edu_youth, income) %>%
  tbl_summary(
    missing = "ifany",
    label = list(
      gender ~ "Gender",
      race_eth ~ "Race / Ethnicity",
      edu_adult ~ "Adult Education",
      edu_youth ~ "Youth Education",
      income ~ "Household Income")) %>%
  bold_labels() %>%
  as_gt()
```

This table describes demographic characteristics for all female participants in our study, from 2011-2018, for a total of 19,848 participants. When available, information on race/ethnicity, education attainment (split by adults >=20, or youth <=19), and household income is reported. 
```{r all_years_table, echo=FALSE, message=FALSE, warning=FALSE}
cat("NHANES Demographics: All Years (2013-2018)")
demo_all_table %>%
  select(gender, age, race_eth, edu_adult, edu_youth, income) %>%
  tbl_summary(
    missing = "ifany",
    label = list(
      gender ~ "Gender",
      age ~ "Age",
      race_eth ~ "Race / Ethnicity",
      edu_adult ~ "Adult Education",
      edu_youth ~ "Youth Education",
      income ~ "Household Income")) %>%
  bold_labels() %>%
  as_gt()
```

